---
title: "Bike Share Exploration"
author: "Partha Sarker"
date: "2023-02-13"
output: html_document
---

```{r}
#load packages
library(tidyverse)
library(janitor)
library(lubridate)
```

```{r}
#import data
df1 <- read.csv("/Users/partha/Documents/R/Bike Share Case Study/t1.csv")
df2 <- read.csv("/Users/partha/Documents/R/Bike Share Case Study/t2.csv")
df3 <- read.csv("/Users/partha/Documents/R/Bike Share Case Study/t3.csv")
df4 <- read.csv("/Users/partha/Documents/R/Bike Share Case Study/t4.csv")
df5 <- read.csv("/Users/partha/Documents/R/Bike Share Case Study/t5.csv")
df6 <- read.csv("/Users/partha/Documents/R/Bike Share Case Study/t6.csv")
df7 <- read.csv("/Users/partha/Documents/R/Bike Share Case Study/t7.csv")
df8 <- read.csv("/Users/partha/Documents/R/Bike Share Case Study/t8.csv")
df9 <- read.csv("/Users/partha/Documents/R/Bike Share Case Study/t9.csv")
df10 <- read.csv("/Users/partha/Documents/R/Bike Share Case Study/t10.csv")
df11 <- read.csv("/Users/partha/Documents/R/Bike Share Case Study/t11.csv")
df12 <- read.csv("/Users/partha/Documents/R/Bike Share Case Study/t12.csv")
```

```{r}
##
## combine 12 data. frames into one (1) data frame
##
bike_rides <- rbind(df1,df2,df3,df4,df5,df6,df7,df8,df9,df10,df11,df12)
bike_rides <- janitor::remove_empty(bike_rides,which = c("cols"))
bike_rides <- janitor::remove_empty(bike_rides, which = c("rows"))

## remove na values
sum(is.na(bike_rides))
bike_rides <- na.omit(bike_rides)
# above ok - do not delete
```

```{r}
## Filtering starting stations with blank names
bike_rides_start_blank <- bike_rides %>% filter (start_station_name =="") %>%
  group_by(ride_id, rideable_type, started_at, ended_at, start_station_name,
           start_station_id, end_station_name, end_station_id, start_lat,
           start_lng, end_lng, member_casual) %>%
  summarise(n=n())

## Filtering starting stations without blank names
bike_rides_start_non_blank <- bike_rides %>% filter (start_station_name !="") %>%
  group_by(ride_id, rideable_type, started_at, ended_at, start_station_name,
           start_station_id, end_station_name, end_station_id, start_lat,
           start_lng, end_lng, member_casual) %>%
  summarise(n=n())

## #Filtering ending stations with blank names
bike_rides_end_blank <- bike_rides %>% filter (end_station_name =="") %>%
  group_by(ride_id, rideable_type, started_at, ended_at, start_station_name,
           start_station_id, end_station_name, end_station_id, start_lat,
           start_lng, end_lng, member_casual) %>%
  summarise(n=n())

## #Filtering ending stations without blank names
bike_rides_end_non_blank <- bike_rides %>% filter (end_station_name !="") %>%
  group_by(ride_id, rideable_type, started_at, ended_at, start_station_name,
           start_station_id, end_station_name, end_station_id, start_lat,
           start_lng, end_lng, member_casual) %>%
  summarise(n=n())

###below chunk necessary?
#merging both start station names
start_data <- data.frame(bike_rides_start_non_blank)
start_data <- rbind(start_data,bike_rides_start_blank)

#merging both end station names
end_data <- data.frame(bike_rides_end_non_blank)
end_data <- rbind(end_data,bike_rides_end_blank)
```

```{r}
#Merging start and end dfs
bike_rides2 <- merge(bike_rides_start_non_blank, bike_rides_end_non_blank, by = "ride_id")
```

```{r}
##
##Convert Date/Time stamp to Date/Time ...
##
bike_rides$Ymd <- as.Date(bike_rides$started_at)
bike_rides$started_at <- 
  lubridate::ymd_hms(bike_rides$started_at)
bike_rides$ended_at <-
  lubridate::ymd_hms(bike_rides$ended_at)

bike_rides$start_hour <-
  lubridate::hour(bike_rides$started_at)
bike_rides$end_hour <- lubridate::hour(bike_rides$ended_at)
```

```{r}

##
##Calculate duration and remove rows with "0" time
##Removed null values from "df"
##
bike_rides$Hours <-
  difftime(bike_rides$ended_at,bike_rides$started_at,units= 
             c("hours"))

bike_rides$Minutes <-
  difftime(bike_rides$ended_at,bike_rides$started_at,units= 
             c("mins"))

df <- bike_rides %>% filter(Hours >0) %>% drop_na() %>%
  select(-ride_id,-end_station_name, -end_station_id,
         -end_station_name)
#fewer obs in "df" data frame - by removing rows with "0" time
#select function to choose only columns of interest
```

Data Log:

10,552 rows with negative trip durations 122,175 rows with missing
Starting station names (and ID) Up to 1291 stations with two different
station_ID Advised by Manager -- ok to ignore rows with trip duration
\<=0 Therefore -- we can ignore rows with missing start station_name,
and also ignore station_ID. And should focus on using the
"start_station_name to perform aggregate functions on date,
start_station_name, member_casual and rideable_type

Cleaning reduced the data to 3,354,514 obs from an original or 3,489,748
obs

```{r}
### Create summary data frame

bikesrides2 <- bike_rides %>% group_by(Weekly =
  floor_date(Ymd,"week"),start_hour) %>%
  summarise(
    Minutes = sum(Minutes),
    Mean = sum(Minutes),
    Median = median(Minutes),
    Max = max(Minutes),
    Count = n()
  ) %>% ungroup()
```

##Plot of Rides By Date \#### Summary Stats: Counts

```{r}
# Summary of Hourly Counts
summary(bikesrides2$Count)
```

-   Count of rides by Hour

```{r}
# Table of Counts by Hour
xtabs(bikesrides2$Count~bikesrides2$start_hour)
```

```{r}
bikesrides2$Monthly <- lubridate::month(bikesrides2$Weekly)

bikesrides2 %>% ggplot() + geom_col(aes(x=Weekly,y=Count,fill=Count)) +
  scale_y_continuous(labels = ,) +
  labs(title = "Count of Rides per Day",
       subtitle = "(Bases on 28 day moving average",
       y = "Average rides per day")
```

```{r}
options(scipen = 999)
bikesrides2 %>% ggplot() + geom_col(aes(x=start_hour,y=Count, fill=Count)) +
  scale_y_continuous(labels = ,) +
  labs(title = "Count of Rides by Hours",
       y = "Rides per Hour")
```

```{r}
### Create summary bikestype data frame
bikestype <- bike_rides %>% group_by(member_casual,rideable_type,Weekly =
  floor_date(Ymd,"week")) %>%
  summarise(
    Minutes = sum(Minutes),
    Mean = sum(Minutes),
    Median = median(Minutes),
    Max = max(Minutes),
    Count = n()
  ) %>% ungroup()
```

-   Count by Bike Type(Total per week)

```{r}
ggplot(bikestype) +
  geom_area(aes(x=Weekly,y=Count,fill=rideable_type)) +
  scale_y_continuous(labels = ,) +
  labs(title = "Count of ride by Bike Type",subtitle = "(for 12 months ending 31-03-2021)")
```

-   Total Ride Minutes by week

```{r}
ggplot(bikestype) + 
  geom_col(aes(x = Weekly,y = Count, fill=member_casual)) +
  scale_y_continuous(labels = ,) +
  labs(title = "Total Ride Minutes by Rider Type")
```

```{r}
bike_rides %>%
  count(start_station_name,sort = TRUE) %>%
  top_n(20) %>% ggplot() +
  geom_col(aes(x=reorder(start_station_name,n),y=n, fill=start_station_name)) +
  coord_flip() +
  labs(title = "Top 20 Start Stations by Ride Count", y="Count of Rides") +
  scale_y_continuous(labels=,)
```

```{r}
# Top 20 start stations for casual riders
bike_rides %>% 
  group_by(start_station_name, member_casual) %>% 
  summarize(number_of_rides = n(), .groups = 'drop') %>% 
  filter(start_station_name != "", member_casual != "member") %>% 
  arrange(-number_of_rides) %>% 
  head(n=20)
```

KEY FINDINGS

-   There are more casual riders on weekends than week days.
-   During the summer months the number of casual riders steadily
    increase between the months of June to August
-   Docked bikes are the most popular and classic bikes are the least
    popular
-   Between 3 PM to 6 PM is the peak time with highest number of rides
    during the day
-   The mean ride length of casual riders is two times longer than
    members
-   The top 3 start stations for casual riders are 1. Streeter Dr. &
    Grand Ave. 2. Lake Shore Dr.& Monroe St. and 3. Millennium Park

Recommendations:

-   Offer attractive discount on memberships during summer months and
    especially weekends to attract casual riders.
-   Marketing campaigns targeting top stations used by casual riders may
    increase the rate of subscription.
-   Form partnership with local businesses around the top 10 stations
    used by casual riders.
-   Carryout marketing campaigns on digital media, cycling apps,
    billboards and local businesses on key cycling routes.
